---
title: "Example workflow for regsplice package"
author: "Lukas M. Weber"
date: "`r doc_date()`"
package: "`r pkg_ver('regsplice')`"
output: BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Example workflow for regsplice package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

\pagebreak



# Introduction

The `regsplice` package implements statistical methods for the detection of differential exon usage (differential splicing) in RNA sequencing (RNA-seq) and exon microarray data sets.

The `regsplice` methods are based on the use of the lasso (L1-regularization) to improve the power of standard generalized linear models, with fast runtimes compared to other leading approaches. The statistical methodology and comparisons of performance with other methods are described in our paper:

> Title of paper and link to bioRxiv preprint when available here.



## Example workflow

This vignette demonstrates an example workflow for the `regsplice` package using a small simulated RNA-seq data set.

There are two options for running `regsplice`: you can run a complete workflow in one step using the wrapper function `regsplice()`; or you can run the individual functions for each step in sequence, which provides additional flexibility and insight into the methodology. Both options are demonstrated below.



## Data set

The data set used for the example workflow consists of exon-level read counts for a subset of 100 genes from a simulated human RNA-seq data set, consisting of 6 biological samples, with 3 samples in each of 2 conditions.

The original data set is from the paper:

> Soneson et al. (2016), *Isoform prefiltering improves performance of count-based methods for analysis of differential transcript usage*, Genome Biology, [available here](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-015-0862-3)

Original data files from this paper, containing the simulated RNA-seq reads (FASTQ and BAM files), are available from ArrayExpress at accession code [E-MTAB-3766](http://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-3766/).

Exon bin counts were generated with the Python counting scripts provided with the [DEXSeq](http://bioconductor.org/packages/release/bioc/html/DEXSeq.html) package, using the option to exclude exons from overlapping genes instead of aggregating them into multi-gene complexes (see Soneson et al. 2016, Supplementary Material).

For this example workflow, we have selected a subset of the first 100 genes from this simulated data set. The exon-level read counts and the true differential splicing status labels for these 100 genes are saved in the text files `vignette_counts.txt` and `vignette_truth.txt` in the `extdata/` directory in the `regsplice` package source code.



## Exon microarray data

The `regsplice` methods are designed to work with both RNA-seq read counts and exon microarray intensities. If you are using exon microarray data, simply provide a matrix or data frame of exon microarray intensities instead of RNA-seq read counts for the `counts` input argument. The name of the argument will still be `counts`, regardless of the type of input data.

The steps in the workflow are the same as shown for RNA-seq data below. See `?regsplice` for the required format for the input matrix or data frame.

Note that exon microarray intensities should be log2-transformed, which can either be done externally (for example with `limma-voom`) or with the `voom_norm = TRUE` argument. In addition, the filtering parameters `filter_n1` and `filter_n2` need to be adjusted carefully when using exon microarray data; filtering may also be disabled with `filter = FALSE`.



## Complete list of user options

For a complete list and description of all available user options, refer to the documentation for the `regsplice()` wrapper function, which can be accessed with `?regsplice` or `help(regsplice)`.




# Workflow

## Load data and create condition vector

Load the vignette example data file, which contains simulated RNA-seq read counts for 100 genes across 6 biological samples. Extract the table of counts and the gene IDs from the raw data.

Then create the `condition` vector, which specifies the experimental conditions or treatment groups for each biological sample.

\vspace{6pt}

```{r}
# load data
file_counts <- system.file("extdata/vignette_counts.txt", package = "regsplice")
data <- read.table(file_counts, header = TRUE, sep = "\t", stringsAsFactors = FALSE)

head(data)

dim(data)

# extract counts and gene IDs from raw data
counts <- data[, 2:7]
gene <- sapply(strsplit(data$exon, ":"), function(s) s[[1]])

head(gene, 6)

# create condition vector
condition <- rep(c("untreated", "treated"), each = 3)

condition
```



## Run workflow with wrapper function

The `regsplice()` wrapper function runs the complete workflow with one command.

The results of a `regsplice` analysis consist of a set of multiple testing adjusted p-values (Benjamini-Hochberg false discovery rates, FDR) quantifying the statistical evidence for differential exon usage (DEU) for each gene. The adjusted p-values are used to rank the genes in the data set according to their evidence for DEU, and an appropriate significance threshold can be used to generate a list of genes with statistically significant evidence for DEU.

The wrapper function also returns gene names, raw p-values, likelihood ratio (LR) test statistics, and degrees of freedom of the LR tests.

The required inputs for the `regsplice()` wrapper function are `counts` (matrix or data frame of RNA-seq read counts or exon microarray intensities), `gene` (vector of gene IDs), and `condition` (vector of experimental conditions for each sample).

See `?regsplice` or `help(regsplice)` for additional details, including other available inputs and options. Note that the progress bar does not display well in the vignette; it will display on a single line on your screen.

\vspace{6pt}

```{r}
library(regsplice)

res <- regsplice(counts, gene, condition)

str(res)
```



## Run workflow with functions for individual steps

Alternatively, you can run the individual functions for each step in the `regsplice` workflow in sequence, which provides additional flexibility and insight into the statistical methodology. The steps are described below.



### Prepare data

The first step consists of pre-processing the input data and preparing it into the format required by other functions in the `regsplice` pipeline. This is done with the `prepare_data()` function.

Inputs are a table of RNA-seq read counts or exon microarray intensities (`counts`) and a vector of gene IDs (`gene`). The vector `gene` must have length equal to the number of rows in `counts`; i.e. one entry for each exon, with repeated entries for multiple exons within the same gene. The repeated entries are used to determine gene length.

The `prepare_data()` function removes exons (rows) with zero counts in all biological samples (columns); splits the count or intensity table into a list of sub-tables (data frames), one for each gene; and removes any remaining single-exon genes. The output is a list of data frames, where each data frame contains the RNA-seq read counts or exon microarray intensities for one gene.

For more details, see `?prepare_data`.

\vspace{6pt}

```{r}
library(regsplice)

Y <- prepare_data(counts, gene)

# length equals the number of genes
length(Y)
```



### Filter low-count exons

Next, we filter low-count exons with `filter_exons()`.

The optional arguments `n1` and `n2` control the amount of filtering. Default values are provided; however these may not be optimal for some experimental designs, so you should consider how much filtering is appropriate for your data set. Any remaining single-exon genes are also removed.

For more details, see `?filter_exons`.

\vspace{6pt}

```{r}
Y <- filter_exons(Y)

# length equals the number of genes
length(Y)
```



### 'voom' weights

In many cases, power to detect differential exon usage can be improved by using exon-level precision weights.

The `voom_weights()` function uses `limma-voom` to calculate exon-level precision weights; as well as an optional log2-counts per million continuous transformation and scale normalization across samples, which in some cases may further improve power.

For more information, see the following paper, which introduced `voom`; or the [limma User's Guide](http://bioconductor.org/packages/release/bioc/html/limma.html) available on Bioconductor.

- Law et al. (2014), *voom: precision weights unlock linear model analysis tools for RNA-seq read counts*, Genome Biology, [available here](http://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29)

By default, `voom_weights()` only returns the weights (see `?voom_weights` for further information). If you wish to use the transformed and normalized data, set the argument `norm = TRUE`.

Note that `voom` assumes that exons (rows) with zero or low counts have already been removed, so this step should be done after filtering.

Exon microarray intensities should be log2-transformed prior to model fitting; so if you are using exon microarray data, either transform the data externally (for example with `voom`), or set the argument `norm = TRUE`.

For more details, see `?voom_weights`.

\vspace{6pt}

```{r}
out_voom <- voom_weights(Y, condition)

weights <- out_voom$weights
```



### Optional: Create design matrices

The function `create_design_matrix()` creates the model design matrix for each gene. This function is called automatically by the model fitting functions, so you do not need to run it directly. Here, we demonstrate how it works for a single gene and show an example design matrix, to provide insight into the statistical methodology.

The design matrix includes main effect terms for each exon and each sample, and interaction terms between the exons and conditions.

Note that the design matrix does not include main effect terms for the conditions, since these are absorbed into the main effect terms for the samples. In addition, the design matrix does not include an intercept column, since it is simpler to let the model fitting functions add an intercept term later.

For more details, see `?create_design_matrix`.

\vspace{6pt}

```{r}
# gene with 3 exons
# 4 biological samples; 2 samples in each of 2 conditions
design_example <- create_design_matrix(condition = rep(c(0, 1), each = 2), n_exons = 3)

design_example
```



### Fit models

There are three model fitting functions: 

- `fit_models_reg()` fits regularized (lasso) models containing an optimal subset of exon:condition interaction terms for each gene. The model fitting procedure penalizes the interaction terms only, so that the main effect terms for exons and samples are always included. This ensures that the null model is nested, allowing likelihood ratio tests to be calculated.

- `fit_models_null()` fits the null models, which do not contain any interaction terms.

- `fit_models_GLM()` fits full GLMs, which contain all exon:condition interaction terms for each gene.

The fitting functions fit models for all genes in the data set. The functions are parallelized using `BiocParallel` for faster runtime. For `fit_models_reg()`, the default number of processor cores is 8, or the maximum available if less than 8. For `fit_models_null()` and `fit_models_GLM()`, the default is one core, since these functions are already extremely fast.

Note that in this example, we have used a single core for `fit_models_reg()`, in order to simplify testing and compilation of this vignette. We have also used `suppressWarnings()` to hide warning messages related to the small number of observations per gene in this data set.

The `weights` argument is optional. If it is not included (default value is `NULL`), exons are weighted equally.

For more details, see `?fit_models_reg`, `?fit_models_null`, or `?fit_models_GLM`.

\vspace{6pt}

```{r}
# fit regularized models
fit_reg <- suppressWarnings(fit_models_reg(Y, condition, weights, n_cores = 1))

# fit null models
fit_null <- fit_models_null(Y, condition, weights)

# fit full GLMs (not required if 'when_null_selected = "ones"' in next step)
fit_GLM <- fit_models_GLM(Y, condition, weights)
```



### Calculate likelihood ratio tests

The function `LR_tests()` calculates likelihood ratio (LR) tests between the fitted models and null models.

If the fitted regularized (lasso) model contains at least one exon:condition interaction term, the LR test compares the lasso model against the nested null model. However, if the lasso model contains zero interaction terms, then the lasso and null models are identical, so the LR test cannot be calculated. The `when_null_selected` argument lets the user choose what to do in these cases: either set p-values equal to 1 (`when_null_selected = "ones"`); or calculate a LR test using the full GLM containing all exon:condition interaction terms (`when_null_selected = "GLM"`), which reduces power due to the larger number of terms, but allows the evidence for differential exon usage among these genes to be distinguished. You can also return NAs for these genes (`when_null_selected = "NA"`).

The default option is `when_null_selected = "ones"`. This simply calls all these genes non-significant, which in most cases is sufficient since we are more interested in genes with strong evidence for differential exon usage. However, if it is important to rank the low-evidence genes in your data set, use the `when_null_selected = GLM` option. If `when_null_selected = "ones"` or `when_null_selected = "NA"`, the full GLM fitted models are not required, so you can set `fit_GLM = NULL` (the default).

The result is a list containing gene names, raw p-values, multiple testing adjusted p-values (Benjamini-Hochberg false discovery rates, FDR), LR test statistics, and degrees of freedom of the LR tests for each gene.

For more details, see `?LR_tests`.

\vspace{6pt}

```{r}
res_no_wrapper <- LR_tests(fit_reg, fit_null)

str(res_no_wrapper)
```



## Summary table of results

After the results have been calculated (either with the wrapper function or the individual functions), the function `summary_table()` can be used to display the results in a more readable format.

The results are displayed as a data frame of the top `n` most highly significant genes, ranked according to either the false discovery rate (FDR) or raw p-values, up to a specified significance threshold.

Set `n = Inf` to display results for all genes up to the specified significance threshold. Set `n = Inf` and `threshold = 1` to display results for all genes in the data set.

Note that we have used `n = 10` for display purposes in this vignette. Many of the p-values and adjusted p-values (FDRs) are exactly equal to zero in this example, due to the strong differential splicing signal in the simulated data set.

For more details, see `?summary_table`.

\vspace{6pt}

```{r}
summary_table(res, n = 10)
```




# Analyze results

Total number of significant genes

Show some plots using external functions



